# Methodology

```{r setup, file = "R/chapter_start.R", include = FALSE}
# a number of commands need to run at the beginning of each chapter. This
# includes loading libraries that I always use, as well as options for 
# displaying numbers and text.
```

```{r setup2, include = FALSE}
# Other packages ------
# These are packages that get used in this chapter but aren't part of the default set.
library(kableExtra)
```

To estimate the impact of Utah's IMT program expansion, we collect data from UDOT and UHP and model the impacts with a linear regression analysis.

## Data

Describe the UHP incident database.

Describe the UDOT incident database.

Describe efforts to clean and filter the data for usable records.

```{r data}
tar_load(ett_data)
```


```{r tab-severity}
#| label: tbl-severity
#| tbl-cap: Descriptive Statistics of Data by Crash Severity
ds_tab <- datasummary_balance(~ type,
                    data = ett_data |> 
                      transmute(
                        type, `IMT responding` = n_imt, `UHP responding` = n_uhp, 
                        `Response time` = rt_imt, `Roadway clearance time` = rct_imt,
                        `Incident clearance time` = ict_imt,
                        `Total lanes on roadway` = lanes_bottle,
                        `Lanes closed for incident` = lanes_closed,
                        `Affected volume` = vol_affect,
                        `Excess travel time` = ett,
                        Year = year,
                        Period = period),
                    output = 'kableExtra'
                    ) 
ds_tab |> 
  kable_styling(position = "center") |> 
  landscape()
```

@tbl-severity presents descriptive statistics of the cleaned incident data by
crash severity. Virtually all of the records are either property damage or
personal injury crashes (approximately half each), with a handful of fatal
crashes. The mean excess travel time for these fatal crashes is approximately
three times higher than for the other more common crash types, though the wide
standard errors and the high degree of skewness make conclusive statements
about this question somewhat difficult.

```{r fig-frequency}
#| label: fig-frequency
#| fig-cap: IMT-responding incidents per day by crash severity.
ett_data |> 
  group_by(date, type) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = n, fill = type)) + 
  geom_bar() + 
  xlab("Incidents per Day") + 
  ylab("Count") +
  theme_bw()
```

@fig-frequency shows a distribution of incidents by day. This might not be the
most relevant plot, but it's the one that I put together.

## Models

We hypothesize that the increase in IMT units decreased response time
as well as decreased excess travel time.  This suggests two models,
$$ 
\log(\mathrm{IMT\ Response\ Time}_i) = X_i\beta
$$
and 
$$ 
\log(\mathrm{Excess\ Travel\ Time}_i) = X_i\beta + \delta\log(\mathrm{IMT\ Response\ Time}_i)
$$
where the index $i$ denotes a single incident, $X$ is a vector of controlling variables --- the number of responding units, the size of the roadways, etc. --- and $\beta$ are estimated coefficients.